---
title: "R Data Visualization"
author: "Azizkhan Myrzakassymov"
date: "2026-04-18"
output: html_document
---

```{r setup, include=FALSE}
# This is a chunk (a code execution block). The first chunk contains
# options for the entire R Markdown document. By setting TRUE/FALSE in
# opts_chunk$set, you apply these settings to every chunk in the document.
# These options include:
# eval    - whether to execute the chunk
# echo    - whether to display the code
# warning - whether to show warnings
# message - whether to show messages
# results - whether to display or hide the results of chunk execution
# (options = 'show' | 'hide')

# The settings below apply to all chunks in the document.
knitr::opts_chunk$set(echo = TRUE)

# If you want apply smth to the exact chunck, write the options in braces at the
# beginning of this chunck, e.g. {r, warning = FALSE }
```

# Markdown Basics here

<https://rmarkdown.rstudio.com/authoring_basics.html>

# For your tables

<https://www.tablesgenerator.com/>

```{r}
library(mlbench)
library(dplyr)
library(ggplot2)
library(ggpubr)
```

```{r}
data(PimaIndiansDiabetes)
pima <- PimaIndiansDiabetes

head(pima)
```

```{r}
# skimr the library for table summary
library(skimr)
skim(pima)
```

```{r}
# the most popular library in R is called ggplot2
library(ggplot2)
```

```{r}
# Let's make our own graph
ggplot(data = pima,
       aes(x = glucose))
# aes is for "aesthetic" 
```

Something clearly went wrong.

In Python’s Matplotlib, you build plots on a figure (like a blank canvas) using axes.

In ggplot2, we’ve created the base figure—now let’s add a plot layer to this “blank canvas.”

For any plot you need:

1.  Data
2.  Aesthetic
3.  Plot type indicator (geom)

How do you choose the right plot type? Start by taking a closer look at your data.

What is the type of your data?

## One numeric value.

### Histogram

```{r}
ggplot(data = pima, aes(x = glucose)) + geom_histogram()
```

### Density plot

```{r}
ggplot(data = pima,
       aes(x = glucose)) + geom_density()
```

### Box plot

```{r}
ggplot(data = pima,
       aes(x = glucose)) + geom_boxplot()
```

## One numerical + one categorical value

### Boxplot

```{r}
ggplot(data = pima,
       aes(x = glucose, y = diabetes)) + geom_boxplot()
```

*Some offtop tips*

```{r}
# Three ways of building a chart
# 1
ggplot(data = pima, aes(x = glucose,
                        y = diabetes)) +
  geom_boxplot()

# 2
pima %>%
  ggplot() +
  geom_boxplot(aes(x = glucose, y = diabetes))

# 3
ggplot() +
  geom_boxplot(data = pima,
               aes(x = glucose, y = diabetes))
```

```{r}
# adding some themes
ggplot() +
  geom_boxplot(data = pima,
               aes(x = glucose, y = diabetes)) + 
  theme_minimal()
```

```{r}
# add some constant values

glucose_mean <- round(mean(pima$glucose), 1)

ggplot(data = pima,
       aes(x = glucose)) + 
  geom_density() +
  geom_vline(aes(xintercept = glucose_mean)) + # adding vertical line
  annotate("text",
           x = glucose_mean+30,
           y = 0.013,
           label = paste0("Mean = ", glucose_mean), color = "red")
```

How to customize the plot:

```{r}
ggplot() + 
  geom_boxplot(data = pima,
               aes(x = glucose, y = diabetes),
               color = 'blue') + 
  theme_minimal() +
  ggtitle('Distribution of a glucose level in diabetes and non-diabetes group') +
  labs(y = "Diabetes", x = 'Glucose level') +
  theme(
    title = element_text(size = 12),
    axis.title.y = element_text(size = 14)) + 
  coord_flip()
                    
```

About colors: [https://www.cookbook-r.com/Graphs/Colors\_(ggplot2)/](https://www.cookbook-r.com/Graphs/Colors_(ggplot2)/){.uri}

## One numerical and two categorical

```{r}

pima <- pima %>%
    mutate(
      diabetes_ch = as.character(diabetes),
      age_group = case_when(
        age < 35 ~ "21-34",
        age >= 35 & age < 50 ~ "35-49",
        age >= 50 ~ "50+"
      )
    )

table(pima$age_group)
```

```{r}
ggplot() +
  geom_boxplot(data = pima,
               aes(x = glucose, y = diabetes, fill = age_group)) + 
  theme_minimal() 
```

## One categorical value

### Barplot

```{r}
ggplot() +
  geom_bar(data = pima,
           aes(x = age_group, fill = diabetes)) +
  theme_minimal()
```

### Density plot

```{r}
ggplot() +
  geom_density(data = pima,
               aes(x = glucose, fill = age_group),
               alpha = 0.5) +
  theme_minimal()
```

```{r}
ggplot() +
  geom_boxplot(data = pima,
               aes(x = glucose, fill = age_group)) +
  theme_minimal()
```

##Two categorical values

###Barplot

```{r}
ggplot() +
  geom_bar(data = pima,
           aes(x = age_group, fill = diabetes),
           position = 'dodge') +
  theme_minimal()
```

## Two numerical values

###Scatter plot

```{r}
pima %>%
  filter(mass != 0, triceps != 0) %>%
  ggplot(aes(x=mass, y=triceps)) +
    geom_point(size = 3, shape = 3) +
  theme_minimal()

```

## Two numerical + one categorical value

```{r}
pima %>%
  filter(mass!=0 & triceps != 0 & pressure !=0) %>%
  ggplot(aes(x = mass, 
             y = triceps,
             color = diabetes)) +
  geom_point(size=2)
```

```{r}
pima %>%
  filter(mass!=0 & triceps != 0) %>%
  ggplot(aes(x = mass, y = triceps)) +
  geom_point(size = 0.5) +
  geom_smooth(method = lm,
              color = 'pink', 
              fill = '#69b3a2',
              se = TRUE) +
  theme_minimal()
```

##Three numerical values

###Scatter plot

```{r}
pima %>%
  filter(mass!=0 & triceps != 0 & pressure !=0) %>%
  ggplot(aes(x = mass, 
             y = triceps,
             color = pressure)) +
  geom_point(size=1) +
  geom_smooth(method = lm,
              color = 'lightgreen',
              fill = 'black',
              se = TRUE,
              alpha = 0.2) +
  theme_minimal()
```

```{r}
pima %>%
  filter(mass!=0 & triceps != 0) %>%
  ggplot(aes(x = mass, 
             y = triceps,
             color = diabetes,
             fill = diabetes,
             group = diabetes)) +
  geom_point(size = 1) +
  geom_smooth(method = lm,
              alpha = 0.2,
              se = TRUE) + 
  theme_minimal()
```

## 3+ numerical values

```{r}
pima %>%
  filter(mass!=0 & triceps != 0 & pressure !=0) %>%
  ggplot() +
  geom_point(
    aes(x = mass,
        y = triceps,
        color = pressure,
        shape = age_group,
        alpha = insulin,
        size = diabetes)) +
  theme_minimal()
```

#Facet graphs

## Two numerical + one categorical value

```{r}
pima %>%
  filter(mass!=0 & triceps != 0) %>%
  ggplot(aes(x = mass, 
             y = triceps,
             color = diabetes,
             fill = diabetes,
             group = diabetes)) +
  geom_point(size=0.5) +
  facet_grid(.~diabetes) +
  theme_minimal()
```

# 2+ graphs on one figure

```{r}
scatter <- pima %>%
  filter(mass != 0 &  triceps != 0) %>%
  ggplot(aes(x=mass, y=triceps)) +
  geom_point(size = 1) +
  theme_minimal()

bar <- 
  ggplot() +
  geom_bar(data = pima,
           aes(x = age_group, fill = diabetes),
           position = 'dodge') +
  theme_minimal()

combine_plot <- ggarrange(scatter, bar,
                          ncol=2, nrow = 1)

combine_plot
```

#Corr analysis

```{r}
library(corrplot)
```

```{r}
pima_clear <- pima %>%
  filter(glucose !=0 & 
           pressure !=0 & 
           triceps !=0 & 
           insulin !=0 & 
           mass !=0 & 
           age !=0) %>%
  select(is.integer | is.numeric)

head(pima_clear)
```

```{r}
pima_cor <- cor(pima_clear)
pima_cor
```

```{r}
corrplot(pima_cor, method = 'number')
```

```{r}
library(corrr)
pima_cor %>%
  network_plot(min_cor = .0)
```

#Heat map

```{r}
pima_aggr <- pima %>%
  group_by(age_group, diabetes) %>%
  summarise(N = n())
```

```{r}
pima_aggr %>%
  ggplot(aes(x = age_group, y = diabetes, fill = N)) +
  geom_tile(color = 'black') +
  geom_text(aes(label = N), color = 'white', size = 4) +
  coord_fixed()
```



# Plot essentials

-   `title` → clearly describes what the plot shows
-   `x`, `y` → axis labels (include units where applicable!)
-   a clean, readable theme (`theme_minimal()` or similar)
-   proper text sizing and formatting

```{r}
ggplot(data = pima, aes(x = diabetes, y = glucose)) +
  geom_boxplot(fill = "lightblue", color = "black") +
  
  labs(
    title = "Distribution of Glucose Levels\nby Diabetes Status",
    x = "Diabetes Status",
    y = "Glucose Level (mg/dL)"
  ) +
  
  theme_minimal() +
  
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    axis.title.x = element_text(size = 10),
    axis.title.y = element_text(size = 10)
  )

```

#Save graphs

```{r}
ggsave(
  plot = combine_plot,
  'combine_plot.png',
  width = 8.5,
  height = 9,
  dpi = 1000
)
```

# Plotly

```{r}
library(plotly)
```

## Boxplot

```{r}
plot_ly(
  pima[pima$insulin !=0,],
  x = ~ insulin,
  color = ~ age_group,
  type = 'box'
)
```

### Scatter plot

```{r}
plot_ly(
  data = pima[(pima$insulin != 0) & (pima$glucose !=0),],
  x = ~insulin,
  y = ~glucose)
```

###Scatter plot + essentials

```{r}
plot_ly(
  data = pima[(pima$insulin != 0) & (pima$glucose !=0),],
  x = ~insulin,
  y = ~glucose,
  color = ~diabetes_ch) %>%
  layout(
    title = 'Glucose level vs insulin level in Pima Dataset',
    yaxis = list(title = 'Glucose',
                 zeroline = FALSE),
    xaxis = list(title = 'Insulin',
                 zeroline = FALSE))
```

###Scatter plot + 3D

```{r}
plot_ly(
  data = pima[(pima$insulin != 0) & (pima$glucose !=0),],
  x = ~insulin,
  y = ~glucose,
  z = ~mass,
  size = 0.5,
  color = ~diabetes_ch) %>%
  layout(
    title = 'Glucose level vs insulin level in Pima Dataset')
```

###Scatter plot + color palettes

R color Cheatsheet: https://www.nceas.ucsb.edu/sites/default/files/2020-04/colorPaletteCheatsheet.pdf 

```{r}
plot_ly(
  data = pima[(pima$insulin != 0) & (pima$glucose !=0),],
  x = ~insulin,
  y = ~glucose,
  color = ~age_group)

```
```{r}
plot_ly(
  data = pima[(pima$insulin != 0) & (pima$glucose !=0),],
  x = ~insulin,
  y = ~glucose,
  color = ~age_group,
  colors = 'Set1')
```

###Another plot essentials

```{r}
plot_ly(
  data = pima[(pima$insulin != 0) & (pima$glucose !=0),],
  x = ~insulin,
  y = ~glucose,
  color = ~diabetes,
  size = ~mass,
  text = ~mass,
  hoverinfo = 'text')
```


