---
title: "Stat criteria"
author: "Myrzakassymov Azizkhan"
date: "2026-04-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# install.packages("mlbench")  
# install.packages("tidyverse")
library(mlbench)
library(tidyverse)
data("PimaIndiansDiabetes")
```

# DATASET

Pima Indians Diabetes dataset — is a widely used medical datast applied to binary classification problems in diabetes. It contains medical indicators of women from the Pima indian people (USA) and serves as a benchmark for machine learning and biostatistics research.


```{r}
pima <- PimaIndiansDiabetes

head(pima)
```


```{r pressure, echo=FALSE}
check_normality <- function(x, var_name = "variable") {
  x_clean <- x[x != 0 & !is.na(x)]

  cat("Normality check for:", var_name, "\n")
  cat("Sample size (after removing zeros):", length(x_clean), "\n\n")
  
  hist(x_clean, main = paste("Histogram of", var_name))
  
  qqnorm(x_clean)
  qqline(x_clean, col = "orange")
  
  test <- shapiro.test(x_clean)
  
  cat("Shapiro-Wilk test p-value:", test$p.value, "\n")
  
  if (test$p.value > 0.05) {
    cat("
        Conclusion: Fail to reject the null hypothesis.\n")
    cat("Data are consistent with a normal distribution.\n")
  } else {
    cat("Conclusion: Reject the null hypothesis.\n")
    cat("Data are not normally distributed.\n")
  }
  
}
```

```{r}
check_normality(pima$pressure, "Total blood pressure")
```

```{r DBP boxplots}
pima_bloods <- pima[pima$pressure != 0 & !is.na(pima$pressure), ]

boxplot(pima_bloods$pressure, 
        main = "Distribution of DBP",
        ylab = "DBP")

print(mean(pima_bloods$pressure))

```



### Hypothesis testing


### Hypothesis statement

Is the diastolic blood pressure of a Pima Indian women sample the same as the normal diastolic blood pressure?

H0: The diastolic blood pressure of a Pima Indian woman sample is the same as the normal diastolic blood pressure

H1: The diastolic blood pressure of a Pima Indian woman sample is not the same as the normal diastolic blood pressure

```{r}
# 0 DBP is a 100% mistake
pima_clean <- pima[pima$pressure != 0 & !is.na(pima$pressure), ]

# what is normal DBP?
DBP = 75

test <- t.test(pima_clean$pressure, mu = DBP)
               
if (test$p.value > 0.05) {
    cat("Conclusion: Fail to reject the null hypothesis.\n")
    cat(paste("The mean is not significantly different from", DBP))
  } else {
    cat("Conclusion: Reject the null hypothesis.\n")
    cat(paste("The mean is significantly different from", DBP))
  }
```

### Conclusion

The diastolic blood pressure of a Pima Indian woman sample is not the same as the normal diastolic blood pressure

## Number of pregnancies

### Hypothesis statement

Is the average number of children of a Pima Indian woman the same as the average number of children of a Kazakhstani woman?

H0: The average number of children of a Pima Indian woman is the same as the average number of children of a Kazakhstani woman?

H1: The average number of children of a Pima Indian woman is not the same as the average number of children of a Kazakhstani woman

### Normality check

H0:

H1:

```{r}
check_normality(pima$pregnant, "Number of pregnancies")
```

### Hypothesis testing

Numerical data -\> Not normal distribution -\> single value -\> Wilcoxon test

```{r}
# 0 pregnancies may not be a mistake

# what is the average number of children for 1 woman in KZ?
avg_children = 2.9

test = wilcox.test(pima_clean$pregnant, mu = avg_children, conf.int = TRUE)

test$conf.int
               
if (test$p.value > 0.05) {
    cat("Conclusion: Fail to reject the null hypothesis.\n")
    cat(paste("The mean is not significantly different from", avg_children))
  } else {
    cat("Conclusion: Reject the null hypothesis.\n")
    cat(paste("The mean is significantly different from", avg_children))
  }

```

### Conclusion:

The average number of children of a Pima Indian woman is not the same as the average number of children of a Kazakhstani woman

Is the any differences in glucose level between diabetes and non-diabetes group?

H0: There is no difference in glucose level between diabetes and non-diabetes group

H1: There is a difference in glucose level between diabetes and non-diabetes group

#### Normality check

```{r}
pima_clean <- pima[pima$diabetes != 0 & !is.na(pima$diabetes), ]

pima$diabetes <- as.factor(pima$diabetes)

pima_pos <- pima[pima$diabetes == "pos", ]
pima_neg <- pima[pima$diabetes == "neg", ]

boxplot(glucose ~ diabetes, data = pima)
```

```{r}
check_normality(pima_pos$glucose, "Glucose level in the diabetes group")
```

```{r}
check_normality(pima_neg$glucose, "Glucose level in the non diabetes group")
```

#### Hypothesis testing

!!! The Mann–Whitney U test and the Wilcoxon rank-sum test are the same test. It's expressed using different statistics.

```{r}
wilcox.test(glucose ~ diabetes, data = pima, conf.int = TRUE)

if (test$p.value > 0.05) {
    cat("Conclusion: Fail to reject the null hypothesis.\n")
    cat(paste("The mean of one group is not significantly different from another"))
  } else {
    cat("Conclusion: Reject the null hypothesis.\n")
    cat(paste("The mean of one group is significantly different from another"))
  }

test$conf.int
```

#### Conclusion

There is a significant differences in glucose level between diabetes and non-diabetes group

### Normal Distribution

#### ATTENTION

We're going to create normally distributed numerical data to illustrate the analysis of normally distributed numerical data between 2 cases

```{r}
set.seed(42)

pima$height <- rnorm(n = nrow(pima), mean = 170, sd = 10)

pima_pos <- pima[pima$diabetes == "pos", ]
pima_neg <- pima[pima$diabetes == "neg", ]
```

#### Hypothesis statement

Is the any differences in women's height between diabetes and non-diabetes group?

H0: There is no difference in women's height between diabetes and non-diabetes group

H1: There is a difference in women's height between diabetes and non-diabetes group

#### Normality check

```{r}
boxplot(height ~ diabetes, data = pima)
```

```{r}
check_normality(pima_pos$height, "Height in the diabetes group")
```

```{r}
check_normality(pima_pos$height, "Height in the non diabetes group")
```

#### Hypothesis testing

Numerical data -\> Normal distribution -\> Independent samples -\> Student test

```{r}
test2 <- t.test(height ~ diabetes, data = pima)

cat("p-value:", test$p.value, "\n")
cat("95% CI:", test$conf.int[1], "to", test$conf.int[2], "\n")

if (test$p.value > 0.05) {
  cat("Conclusion: Fail to reject the null hypothesis.\n")
} else {
  cat("Conclusion: Reject the null hypothesis.\n")
}
```


#### Conclusion

There is a difference in women's height between diabetes and non-diabetes group

## Dependent samples

### Not normal distribution

#### Exposure

Imagine we give metformin to every woman with diabetes from this dataset and than we measures glucise again.

```{r}
metformin_effect <- rnorm(length(pima_pos), mean = 30, sd = 15)

pima_pos$glucose_after_metformin <- pima_pos$glucose - metformin_effect

non_responders <- sample(1:nrow(pima_pos), size = round(0.2 * length(pima_pos)))

pima_pos$glucose_after_metformin[non_responders] <- pima_pos$glucose[non_responders] - rnorm(length(non_responders), mean = 5, sd = 3)
```
#### Hypothesis statement

Is the any differences in glucose level in diabetes group after taking metformin?

H0: There is no difference in glucose level in diabetes group after taking metformin

H1: There is a difference in glucose level in diabetes group after taking metformin

#### Normality check

```{r}
boxplot(glucose_after_metformin ~ diabetes, data = pima_pos)
```

```{r}
check_normality(pima_pos$glucose_after_metformin, "Glicose in the diabetes group after taking metformin")
```

#### Hypothesis testing

Numerical data -\> Not normal distribution -\> dependent samples -\> Mann–Whitney U

!!! The Mann–Whitney U test and the Wilcoxon rank-sum test are the same test. It's expressed using different statistics.

```{r}
wilcox.test(pima_pos$glucose, pima_pos$glucose_after_metformin, paired = TRUE, conf.int = TRUE)

if (test$p.value > 0.05) {
    cat("Conclusion: Fail to reject the null hypothesis.\n")
    cat(paste("The mean of one group is not significantly different from another"))
  } else {
    cat("Conclusion: Reject the null hypothesis.\n")
    cat(paste("The mean of one group is significantly different from another"))
  }

test$conf.int
```

#### Conclusion

There is a significant differences in glucose level in diabetes group before and after taking metformin

Numerical data -\> Not normal distribution -\> dependent samples -\> Mann–Whitney U

!!! The Mann–Whitney U test and the Wilcoxon rank-sum test are the same test. It's expressed using different statistics.


#### in case of normal distribution

```{r}
t.test(pima_pos$glucose, pima_pos$glucose_after_metformin, paired = TRUE, conf.int = TRUE)

if (test$p.value > 0.05) {
    cat("Conclusion: Fail to reject the null hypothesis.\n")
    cat(paste("The mean of one group is not significantly different from another"))
  } else {
    cat("Conclusion: Reject the null hypothesis.\n")
    cat(paste("The mean of one group is significantly different from another"))
  }

test$conf.int
```

# Categorical values

## Independent sample

H0: Age group and diabetes status are independent.

H1: Age group and diabetes status are associated.

```{r}
pima$age_group <- ifelse(pima$age > 40, "older", "younger")

table_data <- table(pima$age_group, pima$diabetes)

chisq.test(table_data)

if (test$p.value > 0.05) {
    cat("Conclusion: Fail to reject the null hypothesis.\n")
  } else {
    cat("Conclusion: Reject the null hypothesis.\n")
  }
```



## Dependent samples

H0: The proportion of women with diabetes before taking metformin does not differ from the proportion of women with diabetes after taking metformin.

H1: The proportion of women with diabetes before taking metformin does differ from the proportion of women with diabetes after taking metformin.

```{r}

metformin_effect <- rnorm(length(pima), mean = 30, sd = 15)

pima$glucose_after_metformin <- pima$glucose - metformin_effect

non_responders <- sample(1:nrow(pima), size = round(0.2 * length(pima)))

pima$glucose_after_metformin[non_responders] <- pima$glucose[non_responders] - rnorm(length(non_responders), mean = 5, sd = 3)

pima$diagnosis_after_metformin <- ifelse(pima$glucose_after_metformin > 140, 'pos', 'neg')

pima$diagnosis_after_metformin <- as.factor(pima$diagnosis_after_metformin)
```
``` {r}
tab <- table(pima$diabetes, pima$diagnosis_after_metformin)

mcnemar.test(tab)

if (test$p.value > 0.05) {
    cat("Conclusion: Fail to reject the null hypothesis.\n")
  } else {
    cat("Conclusion: Reject the null hypothesis.\n")
  }
```

## Single value

Lets say 30% of all Pima people has diabetes. 

H0: the proportion of Pima women with diabets in our sample does not differ from the proportion of diabetes in Pima population 

H1: the proportion of Pima women with diabets in our sample does differ from the proportion of diabetes in Pima population 

```{r}
x <- sum(pima$diabetes == "pos")
n <- nrow(pima)

test = prop.test(x, n, p = 0.3, correct = FALSE)

if (test$p.value > 0.05) {
    cat("Conclusion: Fail to reject the null hypothesis.\n")
  } else {
    cat("Conclusion: Reject the null hypothesis.\n")
  }
```

